A data-driven study of cystic fibrosis treatment: identifying sub-types of physiotherapy style and physical activity using unsupervised clustering

### 2019-08-31 draft.tex
ACT Features

\begin{abstract}
  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment. Little is known about how CYPwCF follow this advice in their everyday lives. Physiotherapy has evolved over the decades without this detailed data. Such data is now being collected as part of Project [anonymous] which is the largest CF physiotherapy trial in the UK to date. The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns. In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms. We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance. We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
\end{abstract}

@@

Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally. CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health. Despite recent progress CF remains progressive and incurable. In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [*******UCL****** ref to this or similar].

To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed. The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [*******UCL****** ref to guideline?]. For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref? This was from a survey that Jackie mentioned]. Physiotherapists also encourage children to have a physically active lifestyle [*******UCL****** ref to why physical activity is good]. Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate [*******UCL****** ref]. The evolution of CF physiotherapy of the decades has not been informed by this important evidence.

Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns. One way to reduce the treatment burden of ACT may be by using gamification. To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography]. Physical activity was also assessed using Fitbit activity trackers.

 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification. This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives. The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment. In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques. These daily sub-types can be tracked over time, as interventions such as gaming are introduced.   
\section{Methods}
\subsection{Trial design and analysis approach}
We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [[*******UCL****** ref to details of trial design,is this sufficient? NREC:18-LO-1038 ]. Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions. The data presented here is from the first 8x months of the trial. Participants are: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, and are voluntarily part of the study. All data is completely de-identified for analysis to researchers.

Our analysis seeks to find patterns in physiotherapy data. Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable. We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal. As such, the data is unlabelled and we use unsupervised learning approach to achieve this. Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].

@@

\subsection{Collaboration, trust and interpretability}
Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2]. To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly. We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months). We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.

@@

\begin{table}[h]
  \caption{Physical activity features}
  \label{fitbit_features}
  \centering
  \begin{tabular}{ |l|l|l| }
    \toprule
    \cmidrule(r){1-2}
    Feature Name & Feature description & Waveform extracted from \\
    \midrule
    userif & 1099 & 0.95 \\
    \bottomrule
    \end{tabular}
\end{table}

>>>>

\begin{abstract}
  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and physical activity as part of their treatment. Little is known about how CYPwCF follow this advice in their everyday lives. Physiotherapy has evolved over the decades without this detailed data. Such data is now being collected as part of Project [anonymous] which is the largest CF physiotherapy trial in the UK to date. The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns. In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features from the waveforms. Using unsupervised learning techniques, We identify meaningful sub-types, with k-means clustering achieving the best performance. We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine. As pertaining to physical activity we identify 4 clusters of steps data and 5 clusters of heartrate data.
\end{abstract}

@@

Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born in the UK. CF is a systemic genetic disorder that mainly affects the respiratory and digestive systems. In the lungs excessive thick sticky mucus causes to a cycle of infection, inflammation and lung damage leading to a deterioration in health. Despite recent progress CF remains progressive and incurable. In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [*******UCL****** ref to this or similar].

To promote clearance of mucus from the lungs, daily physiotherapy including airway clearance techniques (ACT) are prescribed. The physiotherapy exercises involve a series of  exhalations against resistance into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly \cite{Villanuevaj4574} [*******UCL****** ref to guideline?]. For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref? This was from a survey that Jackie mentioned]. Physiotherapists also encourage children to have a physically active lifestyle, this has been shown to reduce the rate of decline in lung function [*******UCL****** ref to why physical activity is good]. Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate with more objective methods not widely available [*******UCL****** ref]. The evolution of CF physiotherapy has not been informed by this important evidence.

Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns. One way to reduce the treatment burden of ACT may be by using gamification. To investigate if gamification can change ACT adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography]. Physical activity patterns were also measured using Fitbit activity trackers.

 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify routine ACT (using bespoke pressure sensors) and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification. This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives. The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment. In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques. These daily sub-types can be tracked over time, as interventions such as gaming are introduced.   
\section{Methods}
\subsection{Trial design and analysis approach}
We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [[*******UCL****** ref to details of trial design,is this sufficient? NREC:18-LO-1038 ]. Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions. The data presented here is from the first 8x months of the trial. Participants are: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, and are voluntarily part of the study. All data is completely de-identified for analysis to researchers.

Our analysis seeks to find patterns in physiotherapy data. Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable. We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal. As such, the data is unlabelled and we use unsupervised learning approach to achieve this. Unsupervised clustering has been used to find sub-types in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].

@@

\subsection{Collaboration, trust and interpretability}
Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included \cite{Alaa2018}. To address the issues that arise from a lack of collaboration \cite{Vayena2018}, \cite{Challen231}, \cite{Char2018}, \cite{Ahmad2018} our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly. We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months). We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.

@@

\begin{table}[h]
  \caption{Physical activity features}
  \label{fitbit_features}
  \centering
  \begin{tabular}{ |l|l|l| }
    \toprule
    \cmidrule(r){1-2}
    \textbf{Feature Name} & \textbf{Feature description} & \textbf{Waveform extracted from} \\
    \midrule
    Active Hours & Number of hours with a step count > 500 & Steps waveform \\
    \bottomrule
    \end{tabular}
\end{table}

\appendix
\section{Appendix : Table of ACT treatment features}

\begin{table}[h]
  \caption{ACT features}
  \label{act_features}
  \centering
  \begin{tabular}{ |l|l| }
    \toprule
    \cmidrule(r){1-2}
    \textbf{Feature Name} & \textbf{Feature description} \\
    \midrule
    Breath Count & Number of breaths the person has made though an ACT  \\ 
     ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
     \midrule
     Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
     Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
     Maximum Breath Amplitude & ~ \\
     Standard Deviation of Breath Amplitude & ~\\
     \midrule
     Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment.\\ 
     Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
     Maximum Break Duration & breath duration in a treatment. \\
     Standard Deviation of Breath Duration & ~\\
     \midrule
     Break Count & Number of breaks (between breaths) in a treatment. \\
     \midrule
     Mean Break Duration & Average duration (in seconds) of breaks between breaths.\\ 
     Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
     Maximum Break Duration & breaks duration in a treatment. \\
     Standard Deviation of Break Duration & ~\\
     \midrule
     Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
     ~ & sets. Set is a group of breaths separated by huffing and coughing activity.\\
     \midrule
     Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
     ~ & Used to filter out "test" device sessions that are not part of a treatment.\\
     \midrule
     Pressures Readings Completeness Score & Data completeness score for the day \\
     ~ & If a treatment has less than 150 pressure readings it's completeness \\
     ~ & score is 0. \\
     \midrule
     Treatment Duration & Treatment duration (in minutes).\\
     \midrule
     Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
     ~ & given day). \\
     \midrule
     Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
     ~ & out of the prescribed amount of breaths. \\ 
     \midrule
     Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
     ~ & out of the prescribed number of treatments for this period. \\
     \midrule
     Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days\\
     ~ & (for a given date) out of the prescribed amount of breaths for this period.\ 
     \midrule
     Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
     ~ & out of the prescribed number of treatments for this period. \\
     \midrule
     Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
     ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
     \midrule
     Cumulative Adherence Score & Cumulative adherence score to date. \\
     ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount.\\
     \midrule
     Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
     \midrule
     Skipped Days Count & Skipped days since the last treatment. \\
     \bottomrule
    \end{tabular}
\end{table}
